---
title: "A Complete Simulation Study with causalsim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Complete Simulation Study with causalsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Evaluating causal estimators requires knowing the answer. `causalsim` gives you
that by making the data-generating process explicit: you specify the structural
model, the package simulates data from it, and you measure how well any estimator
recovers the truth you specified.

This vignette walks through a complete simulation study:

1. Define a DGP with one confounder and a known treatment effect
2. Inspect a draw from it to confirm the structural model
3. Compare a naive (unadjusted) estimator to OLS adjustment
4. Vary sample size with `causalsim_grid()` to study finite-sample precision
5. Vary confounding strength to see where naive estimation breaks down

```{r library}
library(causalsim)
```

## Step 1: Define the DGP

The structural model for this study is:

$$
W \sim N(0,1), \quad
A \mid W \sim \text{Bernoulli}\!\left(\text{logistic}(0.5\,W)\right), \quad
Y = 2A + 0.5W + \varepsilon, \quad
\varepsilon \sim N(0,1)
$$

In `causalsim_dgp()` terms: one standard-normal confounder, a constant effect of
2, moderate propensity confounding, and a moderate baseline shift.

```{r dgp}
dgp <- causalsim_dgp(
  n             = 500,
  n_confounders = 1,
  effect        = 2,
  propensity    = "moderate",
  baseline      = "moderate"
)
dgp
```

The true ATE is exact because `effect` is a scalar — no Monte Carlo approximation
needed.

## Step 2: Inspect a Draw

`causalsim_draw()` simulates one dataset. The columns `.tau` and `.p` are the
individual causal effect and propensity score — diagnostic metadata not available
in real observational data.

```{r draw}
dat <- causalsim_draw(dgp, seed = 1L)
head(dat)
```

With moderate confounding, the treated and control groups differ on the
pre-treatment covariate:

```{r confounding-check}
aggregate(W ~ A, data = dat, FUN = mean)
```

That difference is exactly what makes naive regression biased — any estimator
that ignores `W` will confuse baseline variation with treatment effect.

## Step 3: Define Estimators

Both estimators return a named numeric vector with `estimate`, `ci_lower`, and
`ci_upper`. The `ci_*` fields enable coverage and power metrics in addition to
bias and RMSE.

```{r estimators}
# Naive: regresses Y on A only — omits the confounder
naive_est <- function(data) {
  fit <- lm(Y ~ A, data = data)
  est <- coef(fit)[["A"]]
  se  <- sqrt(vcov(fit)["A", "A"])
  c(estimate = est, ci_lower = est - 1.96 * se, ci_upper = est + 1.96 * se)
}

# OLS: adjusts for the observed confounder W
ols_est <- function(data) {
  fit <- lm(Y ~ A + W, data = data)
  est <- coef(fit)[["A"]]
  se  <- sqrt(vcov(fit)["A", "A"])
  c(estimate = est, ci_lower = est - 1.96 * se, ci_upper = est + 1.96 * se)
}
```

## Step 4: Evaluate Each Estimator

`causalsim_eval()` runs `reps` independent replications and returns bias, RMSE,
coverage, and power with Monte Carlo standard errors.

```{r eval-naive}
eval_naive <- causalsim_eval(dgp, naive_est, reps = 300L, seed = 1L)
eval_naive
```

```{r eval-ols}
eval_ols <- causalsim_eval(dgp, ols_est, reps = 300L, seed = 1L)
eval_ols
```

The naive estimator's bias is substantial: treatment is positively correlated
with $W$, which also raises $Y$ through the baseline, so the unadjusted
coefficient absorbs part of that association. OLS eliminates the bias by
conditioning on $W$. Coverage for the naive estimator is well below the nominal
95 % because confidence intervals are centered on the wrong value.

## Step 5: Vary Sample Size with `causalsim_grid()`

`causalsim_grid()` evaluates an estimator over the Cartesian product of the
supplied parameter values. Here we vary `n` across four levels to track how the
OLS estimator's precision improves with more data.

```{r grid-n}
grid_n <- causalsim_grid(
  dgp       = dgp,
  estimator = ols_est,
  vary      = list(n = c(100L, 250L, 500L, 1000L)),
  reps      = 300L,
  metrics   = c("bias", "rmse"),
  seed      = 1L
)
grid_n
```

RMSE roughly halves as $n$ quadruples — consistent with $\sqrt{n}$-rate
convergence for OLS in a correctly specified model. Bias stays near zero
at every sample size.

## Step 6: Vary Confounding Strength

We can vary the propensity string to study how bias grows as confounding
increases. Because `causalsim_grid()` accepts one estimator at a time, we run
it separately for each and combine the results.

```{r grid-confounding}
conf_levels <- list(propensity = c("low", "moderate", "high"))

grid_naive <- causalsim_grid(dgp, naive_est,
                              vary    = conf_levels,
                              reps    = 300L,
                              metrics = "bias",
                              seed    = 1L)

grid_ols   <- causalsim_grid(dgp, ols_est,
                              vary    = conf_levels,
                              reps    = 300L,
                              metrics = "bias",
                              seed    = 1L)

comparison <- rbind(
  cbind(estimator = "naive", grid_naive$results),
  cbind(estimator = "ols",   grid_ols$results)
)
comparison <- comparison[order(comparison$propensity, comparison$estimator), ]
rownames(comparison) <- NULL
comparison
```

Naive bias grows proportionally with confounding strength. OLS remains near zero
across all three levels because $W$ is observed and included in the model.

---

This three-step workflow — define, evaluate, grid — scales naturally to more
complex settings: heterogeneous effects, multiple confounders, or estimators that
require a different adjustment strategy. See `?causalsim_dgp` for the full
covariate specification API and `?covar` for defining non-normal covariates.
