---
title: "A Complete Simulation Study with causalsim"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{A Complete Simulation Study with causalsim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  out.width  = "100%"
)
```

```{css, echo = FALSE}
.callout {
  background-color: #f0f4f8;
  border-left: 4px solid #3a7ebf;
  padding: 0.75em 1em;
  margin: 1.25em 0;
  border-radius: 0 4px 4px 0;
}

.callout-warn {
  background-color: #fff8e1;
  border-left: 4px solid #f0a500;
  padding: 0.75em 1em;
  margin: 1.25em 0;
  border-radius: 0 4px 4px 0;
}

table {
  width: 100%;
  border-collapse: collapse;
  font-size: 0.9em;
  margin: 1em 0;
}

th {
  background-color: #3a7ebf;
  color: white;
  padding: 0.5em 0.75em;
  text-align: left;
}

td {
  padding: 0.4em 0.75em;
  border-bottom: 1px solid #dde;
}

tr:nth-child(even) td {
  background-color: #f5f7fa;
}
```

## Overview

Evaluating causal estimators requires knowing the answer. `causalsim` gives you
that by making the data-generating process explicit: you specify the structural
model, the package simulates data from it, and you measure how well any estimator
recovers the truth you specified.

### What this vignette covers

| Step | Function | What it does |
|---|---|---|
| 1 | `causalsim_dgp()` | Define the structural model and true ATE |
| 2 | `causalsim_draw()` | Simulate one dataset and inspect it |
| 3 | `causalsim_eval()` | Measure estimator performance over many replications |
| 4 | `causalsim_grid()` | Sweep over sample sizes and confounding levels |

```{r library}
library(causalsim)
```

---

## Step 1: Define the DGP

The structural model for this study is:

$$
W \sim N(0,1), \quad
A \mid W \sim \text{Bernoulli}\!\left(\text{logistic}(0.5\,W)\right), \quad
Y = 2A + 0.5W + \varepsilon, \quad
\varepsilon \sim N(0,1)
$$

In `causalsim_dgp()` terms: one standard-normal confounder, a constant effect of
2, moderate propensity confounding (logistic coefficient 0.5), and a moderate
baseline shift.

```{r dgp}
dgp <- causalsim_dgp(
  n             = 500,
  n_confounders = 1,
  effect        = 2,
  propensity    = "moderate",
  baseline      = "moderate"
)
dgp
```

<div class="callout">
The true ATE is exact when `effect` is a scalar; no Monte Carlo approximation
is needed. For function-valued effects, `causalsim_dgp()` approximates the ATE
via 10,000 Monte Carlo draws at construction time.
</div>

---

## Step 2: Inspect a Draw

`causalsim_draw()` simulates one dataset from the DGP. The columns `.tau` and
`.p` are the individual causal effect and propensity score, diagnostic metadata
that is not available in real observational data.

```{r draw}
dat <- causalsim_draw(dgp, seed = 1L)
head(dat)
```

With moderate confounding, the treated and control groups differ on the
pre-treatment covariate:

```{r confounding-check}
aggregate(W ~ A, data = dat, FUN = mean)
```

<div class="callout-warn">
That difference is exactly what makes naive regression biased. Any estimator
that omits `W` will absorb part of its association with `Y` into the treatment
coefficient.
</div>

---

## Step 3: Define Estimators

An estimator is any function that accepts the data frame returned by
`causalsim_draw()` and returns a named numeric vector. The `ci_lower` and
`ci_upper` fields are optional but enable coverage and power metrics.

```{r estimators}
# Naive: regresses Y on A only, omits the confounder
naive_est <- function(data) {
  fit <- lm(Y ~ A, data = data)
  est <- coef(fit)[["A"]]
  se  <- sqrt(vcov(fit)["A", "A"])
  c(estimate = est, ci_lower = est - 1.96 * se, ci_upper = est + 1.96 * se)
}

# OLS: adjusts for the observed confounder W
ols_est <- function(data) {
  fit <- lm(Y ~ A + W, data = data)
  est <- coef(fit)[["A"]]
  se  <- sqrt(vcov(fit)["A", "A"])
  c(estimate = est, ci_lower = est - 1.96 * se, ci_upper = est + 1.96 * se)
}
```

Named lists are also accepted, so the following is equivalent:

```{r estimator-list, eval = FALSE}
ols_est <- function(data) {
  fit <- lm(Y ~ A + W, data = data)
  ci  <- confint(fit)["A", ]
  list(
    estimate = coef(fit)["A"],
    ci_lower = ci[1],
    ci_upper = ci[2]
  )
}
```

---

## Step 4: Evaluate Each Estimator

`causalsim_eval()` runs `reps` independent replications and returns a tidy
summary of bias, RMSE, coverage, and power with Monte Carlo standard errors.

```{r eval-naive}
eval_naive <- causalsim_eval(dgp, naive_est, reps = 300L, seed = 1L)
eval_naive
```

```{r eval-ols}
eval_ols <- causalsim_eval(dgp, ols_est, reps = 300L, seed = 1L)
eval_ols
```

The naive estimator's bias is substantial: treatment is positively correlated
with $W$, which also raises $Y$ through the baseline, so the unadjusted
coefficient absorbs part of that association. OLS eliminates the bias by
conditioning on $W$. Coverage for the naive estimator falls well below the
nominal 95% because the confidence intervals are centered on the wrong value.

`summary()` adds the full distribution of per-replication estimates, and
`plot()` shows it as a histogram:

```{r summary-ols}
summary(eval_ols)
```

```{r plot-ols, fig.cap = "Distribution of OLS estimates over 300 replications. Solid line: true ATE. Dashed line: mean estimate."}
plot(eval_ols)
```

---

## Step 5: Vary Sample Size with `causalsim_grid()`

`causalsim_grid()` evaluates an estimator over the Cartesian product of the
supplied parameter values, returning a tidy data frame of metrics for each cell.
Here we vary `n` across four levels to track how the OLS estimator's precision
improves with more data.

```{r grid-n}
grid_n <- causalsim_grid(
  dgp       = dgp,
  estimator = ols_est,
  vary      = list(n = c(100L, 250L, 500L, 1000L)),
  reps      = 300L,
  metrics   = c("bias", "rmse"),
  seed      = 1L
)
grid_n
```

RMSE roughly halves as $n$ quadruples, consistent with $\sqrt{n}$-rate
convergence for OLS in a correctly specified model. Bias stays near zero at
every sample size.

---

## Step 6: Vary Confounding Strength

Varying the `propensity` preset shows how bias scales with confounding. Because
`causalsim_grid()` accepts one estimator at a time, we run it separately and
combine the results.

```{r grid-confounding}
conf_levels <- list(propensity = c("low", "moderate", "high"))

grid_naive <- causalsim_grid(dgp, naive_est,
                             vary = conf_levels,
                             reps = 300L,
                             metrics = "bias",
                             seed = 1L)

grid_ols <- causalsim_grid(dgp, ols_est,
                           vary = conf_levels,
                           reps = 300L,
                           metrics = "bias",
                           seed = 1L)

comparison <- rbind(
  cbind(estimator = "naive", grid_naive$results),
  cbind(estimator = "ols",   grid_ols$results)
)
comparison <- comparison[order(comparison$propensity, comparison$estimator), ]
rownames(comparison) <- NULL
comparison
```

Naive bias grows proportionally with confounding strength. OLS remains near zero
across all three levels because $W$ is observed and included in the model.

---

## Where to go next

This workflow (define, evaluate, grid) scales to more complex settings. A few
directions:

| Goal | How |
|---|---|
| Heterogeneous effects | Pass a function to `effect` in `causalsim_dgp()` |
| Non-normal covariates | Use `covar("binary")` or `covar("uniform")` in `covariates` |
| Multiple confounders | Set `n_confounders = 3` or pass named `covariates` |
| Custom covariate structure | Mix `n_confounders` with explicit `covariates = list(...)` |

See `?causalsim_dgp` and `?covar` for the full API.

---

## Session info

```{r session-info}
sessionInfo()
```
